Optical Music Recognition for LilyPond File Generation

### 2024-05-09 papers/cs444/paper-cs444.tex
paper progress, lilypond explanation

\usepackage{graphicx}
% \graphicspath{{figures/}}
%\usepackage{multicol}
\usepackage{multirow}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{framed}

@@

\title[omr-lilypond-midi]{Optical Music Recognition for MIDI-LilyPond File Generation}
\author{Evan Matthews}
\email{evanmm3@illinois.edu}
%\orcid{1234-5678-9012}
\affiliation{%
	\institution{University of Illinois Urbana-Champaign}
	\country{}
}

@@

\section{Introduction}
\todo{Introduction}
~\cite{oord2016wavenet,contreras2023omrcnn,contreras2023omrpiano,mayer2024practical,andrea2021note}

@@

\subsection{Lilypond, MIDI}

\section{Experiment}
\subsection{Key resources}
The Image-to-MIDI convolutional neural network will be built from scratch while referencing promising results from OMR and image-to-text research, (should the CNN need fine-tuning or other enhancement). Currently, the model dataset is the complete Bach Midi Index and will be narrowed down to fit additional constraints: two staves, single instrument, rhythmic quantization at the 16th or 32nd- note level. Model testing will focus primarily on self-trained work, but I am interested in comparing against pre-trained models if I have extra time. For hardware, training will take place on a Linux GPU cluster, courtesy of Paris Smaragdis and the UIUC-CS Audio Computing Lab. Programming work is using a combination of Python with Pytorch for the CNN, and Bash scripts with Lilypond to prepare and normalize the dataset.

>>>>

\usepackage{graphicx}
\usepackage{listings}
\usepackage{multirow}
\usepackage{balance}
\usepackage{amsmath}
\usepackage{framed}
\usepackage{subfigure}

@@

\title[omr-lilypond-midi]{Optical Music Recognition for LilyPond File Generation}
\author{Evan Matthews}
\email{evanmm3@illinois.edu}
%\orcid{1234-5678-9012}
\affiliation{%
	\institution{University of Illinois Urbana-Champaign}
	\country{}
}

@@

\section{Introduction}
\todo{Introduction}
\cite{oord2016wavenet,contreras2023omrcnn,contreras2023omrpiano,mayer2024practical,andrea2021note}

@@

\subsection{Lilypond}
Lilypond was chosen for dataset file representation as it maintains accurate translation between digital (MIDI) and visual (image) notation systems~\cite{lilypond}.
In particular, translation from Lilypond to MIDI is trivial, and Lilypond provides a more intuitive representation of musical information compared to raw bytes of MIDI data.
This notation language consists of multiple layers to separate score components:

\begin{itemize}
		\item \textbf{Document level:} components related to page layout and high-level musical details, (number of instruments/tracks, score engraving information).
		\item \textbf{Music level:} components related to low-level musical details, (notes, rhythms, durations, key/time signatures, tempo).
\end{itemize}

At the document level, \todo{document}

\begin{figure*}
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/lilypond-snippet.png}
		% \caption{Rendered Lilypond (.ily) file with code}
		\label{fig:lilypond-snippet}
	\end{subfigure}
	~
	\begin{subfigure}
		\centering
		\includegraphics[width = .8\linewidth]{./figures/lilypond-code.png}
		\label{fig:lilypond-code}
	\end{subfigure}
	\caption{Lilypond (.ily) code and rendering, (time signature = 4/4)}
	\label{fig:lilypond-example}
\end{figure*}

At the music level, rhythmic musical symbols are notated according to their pitches and relative time duration.
pitches are all characters ${a, b, c, d, e, f, g, r}$, where $r$ is a "rest" meaning no pitch occurs.
To ascend or descend in pitch beyond a single musical "octave," the $'$ and $,$ characters are appended to indicate one or more octave ascendings or descendings, respectively.
Pitches are also proceeded by a rhythmic value representing its duration in time.
These values are typically powers of two, (but can technically be any positive floating-point value), and they dictate how long a pitch is played with respect to the piece's time signature.
For example, a score with time signature = 2/4 indicates that each measure has two beats, and each beat is the length of a quarter note (hard-coded in the denominator).
In this case, the line $"c4 d4"$ would represent two quarter (4) notes or a single measure in the provided time signature.
Additionally, notes without defined durational values take on the previous note's duration in a line, so a series of equal-duration notes is represented by one durational value.
Finally, separations between measures are made with $"| \% n"$, marking the end of measure $n$.
Figure~\ref{fig:lilypond-example} compares the rendering of a single staff of music with its relative Lilypond notation .

% \begin{lstlisting}
% trackBchannelB = \relative c {
% 	r16 c' d e f d e c g'8 c b32 a b16 c8 
% 	| % 2
% 	d16 g, a b c a b g d'8 g f32 e f16 g8 
% 	| % 3
% }
% \end{lstlisting}

\section{Experiment}
\subsection{Key resources}
\todo{Remove this section?}
The Image-to-MIDI convolutional neural network will be built from scratch while referencing promising results from OMR and image-to-text research, (should the CNN need fine-tuning or other enhancement). Currently, the model dataset is the complete Bach Midi Index and will be narrowed down to fit additional constraints: two staves, single instrument, rhythmic quantization at the 16th or 32nd- note level. Model testing will focus primarily on self-trained work, but I am interested in comparing against pre-trained models if I have extra time. For hardware, training will take place on a Linux GPU cluster, courtesy of Paris Smaragdis and the UIUC-CS Audio Computing Lab. Programming work is using a combination of Python with Pytorch for the CNN, and Bash scripts with Lilypond to prepare and normalize the dataset.
